Thermal reactionomes reveal adaptive responses to thermal extremes in warm

### 2014-10-09 doc/ApTranscriptome_manuscript.Rmd
started edits to Discussion

---
title: Thermal reactionomes reveal adaptive responses to thermal
  extremes in warm and cold-climate ant species
author: ' '
csl: ecology.csl
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document: null
  word_document: default
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

@@

Following the genetic assimilation hypothesis that adaptation occurs via changes from plastically induced to constitutive expression, we predicted a tradeoff between constitutive expression levels and inducibility of a transcript. In contrast to this prediction, constitutive expression levels at 19.5°C were higher in *A. carolinensis* than *A. picea* in nearly all comparisons. The exception to this was for genes that had **High** or **Bimodal** expression in *A. picea* and were not responsive in *A. carolinensis*. Rather than being constitutively elevated as predicted, these genes were not expressed (...mean TPM...) in *A. carolinensis*. ...gene ontology terms...genes recruited to thermal tolerance... Consistent with our results but at a much shorter time-scale, no evidence for genetic assimilation at the expression level was found in an experimental evolution study with *Caenorhabditis remanei* selected for 10 generations to increased heat tolerance, even though there was a substantial increase in heat tolerance and loss of plasticity in the evolved populations [@sikkink2014]. In this study, the genetic assimilation at the phenotypic level was caused by a shift in the threshold for temperature response, similar to what we observed here (Fig. 2).

While our results did not the genetic assimilation hypothesis, the observation of higher constitutive expression of thermally-responsive transcripts is consistent with the prediction that the warm-climate species *A. carolinensis* would rely more upon "frontloading" than the cool-climate *A. picea*. Moreover, *A. carolinensis* shut-down expression of more transcripts (Fig. 2) at thermal extremes, and had greater inertia of expression to changes in temperature (Fig. 2) than did *A. picea*. In constrast, the cool-climate species *A. picea* activated more transcripts in response to both thermal extremes, including enrichment of biological processes related to metabolism, transport and regulation. One possible explanation for this difference in responses is that *A. picea* mounts a classical cellular stress response [@kultz2005; @richter2010], requiring increased metabolism, while *A. carolinensis* relies upon constitutive defenses [@barshis2013], proteome stability [@ghosh2010] or behavior [@sunday2014] to tolerate thermal stress. Consistent with this hypothesis, we found that of the 7 thermally responsive heat shock proteins, all 7 increased expression in response to thermal extremes in *A. picea*, but only 3 do in *A. carolinensis*. In contrast, by shutting-down molecular activity, must rely upon other mechanisms such as behavior or increased proteome stability to tolerate thermal stress. These differences are in line with the expectations if *A. picea*, a northern species with fewer good foraging days, must forage in suboptimal conditions whereas *A. carolinensis* can, with a longer season, afford to restrain from foraging in suboptimal conditions. In fact, restrained foraging under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].

@@

>>>>

---
title: Thermal reactionomes reveal adaptive responses to thermal extremes in warm
  and cold-climate ant species
author: ' '
csl: ecology.csl
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document: null
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

Under the *enhanced defense* hypothesis, we predicted that the cool-climate species *A. picea* would induce more genes in response to low temperatures than *A. carolinensis*, while *A. carolinensis* would induce more genes in response to high temperatures. This prediction was only true for *A. picea*, which induced the same genes and then recruited additional genes in response to both low and high temperatures (Fig. 1). In addition, we predicted that upregulation should occur in response to less extreme cold and high temperatures for *A. picea* and *A. carolinensis*, respectively. In contrast, we found no difference between species in the mean temperature of expression response for cold temperatures (Fig. 2A,C), and that higher temperatures were required to induce the expression response for *A. carolinensis* than *A. picea* for cold temperatures. While conflicting with the *enhanced defense* hypothesis, this is consistent with the *tolerance* hypothesis for *A. carolinensis*.

We also explored the *genetic assimilation* hypothesis that exposure to more frequent stressors would select for a shift from inducible to constitutive expression of stress-response genes. In contrast to this prediction, constitutive expression levels at 19.5°C were higher in *A. carolinensis* than *A. picea* in nearly all comparisons. The exception to this was for genes that had **High** or **Bimodal** expression in *A. picea* and were not responsive in *A. carolinensis*. Rather than being constitutively elevated as predicted, these genes were not expressed (...mean TPM...) in *A. carolinensis*. ...gene ontology terms...genes recruited to thermal tolerance... Consistent with our results but at a much shorter time-scale, no evidence for genetic assimilation at the expression level was found in an experimental evolution study with *Caenorhabditis remanei* selected for 10 generations to increased heat tolerance, even though there was a substantial increase in heat tolerance and loss of plasticity in the evolved populations [@sikkink2014]. In this study, the genetic assimilation at the phenotypic level was caused by a shift in the threshold for temperature response, similar to what we observed here (Fig. 2).

While our results did not the genetic assimilation hypothesis, the observation of higher constitutive expression of thermally-responsive transcripts is again consistent with the *tolerance* hypothesis for *A. carolinensis*.

Thus, we find evidence that these two species have responded to temperature using distinct mechanisms. *A. picea* has enhanced its thermal response, while *A. carolinensis* tolerates thermal stress...

----------------------------------------------------

with the prediction that the warm-climate species *A. carolinensis* would rely more upon "frontloading" than the cool-climate *A. picea*. Moreover, *A. carolinensis* shut-down expression of more transcripts (Fig. 2) at thermal extremes, and had greater inertia of expression to changes in temperature (Fig. 2) than did *A. picea*. In constrast, the cool-climate species *A. picea* activated more transcripts in response to both thermal extremes, including enrichment of biological processes related to metabolism, transport and regulation. One possible explanation for this difference in responses is that *A. picea* mounts a classical cellular stress response [@kultz2005; @richter2010], requiring increased metabolism, while *A. carolinensis* relies upon constitutive defenses [@barshis2013], proteome stability [@ghosh2010] or behavior [@sunday2014] to tolerate thermal stress. Consistent with this hypothesis, we found that of the 7 thermally responsive heat shock proteins, all 7 increased expression in response to thermal extremes in *A. picea*, but only 3 do in *A. carolinensis*. In contrast, by shutting-down molecular activity, must rely upon other mechanisms such as behavior or increased proteome stability to tolerate thermal stress. These differences are in line with the expectations if *A. picea*, a northern species with fewer good foraging days, must forage in suboptimal conditions whereas *A. carolinensis* can, with a longer season, afford to restrain from foraging in suboptimal conditions. In fact, restrained foraging under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].

@@

Following the genetic assimilation hypothesis that adaptation occurs via changes from plastically induced to constitutive expression, we predicted a tradeoff between constitutive expression levels and inducibility of a transcript.

@@

CTmax and CTmin hypothesized to be genetically correlated, but not evident in terms of gene expression.  Only 15-20% of genes upregulated in response to temp are bimodal (ie generalized stress response), and although upper activation thresholds differed significantly btwn species, lower thresholds were virtually identical.  Implications . . .
